---
title: "Trig substitutions"
format:
  html:
    code-fold: true
    toc: true
    html-math-method: katex
---

Our motivation for this section is the following integral

$$
\int \frac{\sqrt{25x^2 - 4}}{x} \, dx
$$

This integral is very difficult to solve with the previous techniques. The trick comes from the substitution

$$
x = \frac{2}{5}\sec \theta
$$

Which after a fair amount of algebra and calculus gives us

$$
\int \frac{\sqrt{25x^2 - 4}}{x} \, dx = \sqrt{25x^2 - 4} + 2\arccos \left(\frac{2}{5x}\right) + C
$$

Solving these integrals relies on a method known as [**trigonometric substitutions**]{style="color: blue"}.

## Method

First, lets explain how we solved the previous integral.

::: {.callout-warning icon="false" collapse="false"}
## Example

We have the following integral, which is difficult to solve using previous techniques.

$$
\int \frac{\sqrt{25x^2 - 4}}{x} \, dx
$$

Using the substitution

$$
x = \frac{2}{5}\sec \theta
$$

We can reduce the integral and get a solution in terms of $\theta$.

```{=tex}
\begin{array}{ll}

& = \displaystyle \int \frac{\sqrt{25(\frac{2}{5}\sec \theta)^2 - 4}}{\frac{2}{5}\sec \theta} \, (\frac{2}{5}\tan \theta \sec \theta \, d\theta)\\
& =  \displaystyle\int \sqrt{4\sec^2 \theta - 4} \, \tan \theta \, d\theta\\
& = \displaystyle2\int \sqrt{\sec^2 \theta - 1} \, \tan \theta \, d\theta = 2\int\tan^2\theta \, d\theta\\
& = 2\left(\tan \theta + \theta\right) + C
\end{array}
```
Now we only need to undo our substitution. Notice that

$$
\sec \theta = \frac{\text{hyp}}{\text{adj}} = \frac{5x}{2}
$$

This can be visualized in the triangle.

```{r, engine = 'tikz', fig.width=3, fig.height=3}
#| fig-align: 'center'

\usetikzlibrary{graphs,graphs.standard,angles,quotes}

\begin{tikzpicture}

\draw[fill=black] (0,0);
\draw[fill=black] (3,0);
\draw[fill=black] (3,2);
\draw[line width=1pt] (0,0) -- (3,0) -- (3,2) -- (0,0);
\node at (1.75,-.25) {$2$};
\node at (1.2,1.25) {$5x$};
\node at (3.9,1) {$\sqrt{25x^2 - 4}$};

\coordinate (origo) at (0,0);
\coordinate (pivot) at (3,2);
\coordinate (mary) at (3,0);

\pic[draw, -, "$\theta$", angle eccentricity=1.5]{angle = mary--origo--pivot};
\end{tikzpicture}
```

From here can see that
$$
\tan \theta = \frac{\sqrt{25x^2-4}}{2}
$$

$$
\theta = \arccos \left(\frac{2}{5x}\right)
$$

Giving us our final answer

$$
\int \frac{\sqrt{25x^2 - 4}}{x} \, dx = \sqrt{25x^2-4} + 2\arccos \left(\frac{2}{5x}\right) + C
$$
:::

When faced with an integral containing a radical expression in the form $\sqrt{\pm(bx + c)^2 \pm a}$, we can usually find some subtitution for $x$ which simplifies the expression into a singular trigonometric identity. This is achievable due to the following three identities.

```{=tex}
\begin{array}{ll}
1 - \sin^2x &=& \cos^2 x\\ 
\tan^2x + 1 &=& \sec^2 x\\
\sec^2x - 1 &=& \tan^2 x 
\end{array}
```

Lets see how these identites allow us to reduce certain radicals.

::: {.callout-tip icon="false" collapse="true"}
## Trig substitutions

Lets examine the three cases where trig identities allow us to simplify a radical expression.

**Case I:** $\sqrt{a - (bx + c)^2}$

Notice how this case is a constant, subtracted by a squared variable term. This resembles our first expression, $1 - \sin^2x = \cos^2 x$. We want to get the expression in these terms so we can reduce the radicle. Notice by making the substitution

$$
(bx + c) = \sqrt{a}\sin \theta
$$

We get the equation

$$
\sqrt{a - (\sqrt{a}\sin\theta)^2} = \sqrt{a}\sqrt{1-\sin^2\theta} = \sqrt{a}\cos\theta
$$

We have now elimated the variable from the radical.

**Case II:** $\sqrt{(bx + c)^2 + a}$

This case resembles our identity $\tan^2x + 1 = \sec^2 x$. So lets try the substitution

$$
(bx + c) = \sqrt{a}\tan \theta
$$

Which gives us

$$
\sqrt{(\sqrt{a}\tan \theta)^2 + a} = \sqrt{a}\sqrt{\tan^2 \theta + 1} = \sqrt{a}\sec\theta
$$

We have now elimated the variable from the radical.

**Case III:** $\sqrt{(bx + c)^2 - a}$

This case resembles $\sec^2x - 1$. Using the substitution

$$
(bx + c) = \sqrt{a}\sec \theta
$$

We get the equation

$$
\sqrt{(\sqrt{a}\sec \theta)^2 - a} = \sqrt{a}\sqrt{\sec^2\theta - 1} = \sqrt{a}\tan\theta
$$

We have now elimated the variable from the radical.

:::

For ease of use, the following table shows what substitution to make depending on the case.

```{=tex}
\begin{array}{ll}
\text{Radical expression} \hspace{0.25in} & \text{Substitution}\\
&\\
\sqrt{a - (bx + c)^2} \hspace{0.25in} & (bx + c) = \sqrt{a}\sin\theta\\
&\\
\sqrt{(bx + c)^2 + a} \hspace{0.25in} & (bx + c) = \sqrt{a}\tan\theta\\
&\\
\sqrt{(bx + c)^2 - a} \hspace{0.25in} & (bx + c) = \sqrt{a}\sec\theta
\end{array}
```
## Examples

::: {.callout-warning icon="false"}
## Problem

Evaluate the following integral

$$
\int \frac{\sqrt{9 - x^2}}{x^2}\,dx
$$
:::

::: {.callout-tip icon="false" collapse="true"}
## Solution

Notice that the radicle expression resembles the third row of our table. If we make the substitution

$$
x = 3\sin \theta\\
dx = 3\cos\theta\,d\theta
$$

We get

$$
\int \frac{\sqrt{9 - x^2}}{x^2}\,dx = \int\frac{3\sqrt{1-\sin^2\theta}}{9\sin^2\theta}(3\cos\theta)\,d\theta = \int\frac{\cos^2\theta}{\sin^2\theta} d\theta
$$

Which evaulates to
$$
\int\frac{\cos^2\theta}{\sin^2\theta} d\theta = -\cot\theta - \theta + C
$$

This however, is only half the problem. We still need to convert the equation back to its orginal variable. Once again, the method of drawing a triangle will help to simplify the final expression.

Note that $\sin \theta = \displaystyle\frac{\text{opp}}{\text{hyp}} = \displaystyle\frac{x}{3}$

```{r, engine = 'tikz', fig.width=3, fig.height=3}
#| fig-align: 'center'

\usetikzlibrary{graphs,graphs.standard,angles,quotes}

\begin{tikzpicture}

\draw[fill=black] (0,0);
\draw[fill=black] (3,0);
\draw[fill=black] (3,2);
\draw[line width=1pt] (0,0) -- (3,0) -- (3,2) -- (0,0);
\node at (1.75,-.3) {$\sqrt{9 -x^2}$};
\node at (1.2,1.25) {$3$};
\node at (3.2,1) {$x$};

\coordinate (origo) at (0,0);
\coordinate (pivot) at (3,2);
\coordinate (mary) at (3,0);

\pic[draw, -, "$\theta$", angle eccentricity=1.5]{angle = mary--origo--pivot};
\end{tikzpicture}
```

So

$$
\cot \theta = \frac{\text{adj}}{\text{opp}} = \frac{\sqrt{9 -x^2}}{x}
$$

Giving us a final answer of 

$$
\int \frac{\sqrt{9 - x^2}}{x^2}\,dx = \frac{-\sqrt{9-x^2}}{x} - \arcsin\left(\frac{x}{3}\right) + C
$$

:::